---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Set locale so that names of weekdays appear in English,
```{r, echo = TRUE, results = 'hide'}
Sys.setlocale("LC_TIME", "C")
```
and load the neccessary libraries.
```{r, echo = TRUE, results = 'hide'}
library(data.table)
library(ggplot2)
```


## Loading and preprocessing the data
Extract the 'activity.csv' file from the zipfile
```{r, echo = TRUE, results = 'hide'}
unzip(zipfile = 'activity.zip')
```
Read data into a data table and convert the date into POSIXct
```{r, echo = TRUE, results = 'hide'}
activityData <- fread('activity.csv')
activityData[, date := as.POSIXct(activityData[, date], format="%Y-%m-%d")]
```


## What is mean total number of steps taken per day?
Aggregate the total number of steps per date into the variable 'totalSteps',
plot a histogram and calculate mean and median of steps per day.
```{r meanSteps, echo = TRUE}
stepsPerDay <- activityData[, list(totalSteps = sum(steps)), by = date]
hist(stepsPerDay$totalSteps, 
     col  = 'Red',
     main = 'Histogram of total steps per day',
     xlab = 'Number of steps',
     ylab = 'Days'
)
meanStepsPerDay   <- as.integer(stepsPerDay[, mean(totalSteps, na.rm = TRUE)])
medianStepsPerDay <- stepsPerDay[, median(totalSteps, na.rm = TRUE)]
```
Mean number of steps per day: `r meanStepsPerDay`.   
Median number of steps per day: `r medianStepsPerDay`.

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
